15yrsChange

### 2024-04-05 doc/15yrsChange.Rmd
discussion: edited another paragraph, plus a bit of rearrangement

The observed trends in AGB loss, including increasing AWM and decreasing AWP, will not necessarily be maintained as continuous directional trends in the future.
The top driver of increasing AWM has been EAB-driven mortality of *Fraxinus spp.*, but *Fraxinus* AWM has started declining as most individuals ≥10 cm have already died (Fig. `r mortality_by_genera`).
It is possible that increasing AWM will be maintained by trends of increasing oak mortality or by other non-indigenous insects or pathogen, which have arrived in the region in multiple waves over the past century [@anderson-teixeira_longterm_2021]. 
While it is also possible that AWM will return to its previously low level as EAB recedes, we consider it very likely that oak decline, additional non-indigenous insects or pathogen species (e.g., beech leaf disease), and climate stressors will continue to elevate tree mortality at this site in the coming decades.
Similarly, the observed long-term declines at this site in AWP (Fig. `r loss_gain_net`b) and individual tree growth [@anderson-teixeira_joint_2022] remain poorly understood, hindering prediction of their future trends.
Overall, there remains large uncertainty as to the future trend in AGB at this site, but our analysis makes it clear that these trends have been and will continue to be substantially impacted by nuisance species.

Despite inherent limitations in extrapolating findings from a single site, we find robust evidence suggesting a broader applicability of our results. 
Mixed-hardwood forests, akin to our study site, are typical of the mid-Atlantic region. The persistently elevated deer populations documented at this site are reported throughout eastern US forests.
Furthermore, the widespread occurrence of vulnerable species (*e.g.* *Fraxinus* spp.) and continued spread of non-indigenous nuisance species (*e.g.* emerald ash borer, EAB) reinforces the relevance of our findings to the broader regional context. EAB consistently induces near-total mortality in *Fraxinus* spp. once infected, to such an extent that five species in North America (including *Fraxinus americana*) are now critically endangered (Knight et al, 2013, IUCN Red list). The ongoing emergence of new threats endangers other important canopy species: beech leaf disease, a non-indigenous pathogen lethal to *Fagus grandifolia* and first detected in 2012, is rapidly spreading throughout eastern US deciduous forests [@reed_distribution_2022]. Another non-indigenous pathogen, Sudden Oak Death, is currently confined to the Western US but is causing widespread mortality in multiple Oak species and poses huge risks if it were to successfully spread to this region [@conrad_threat_2020].

>>>>

The observed trends in AGB loss, including increasing AWM and decreasing AWP, will not necessarily be maintained as continuous directional trends in the future.
The top driver of increasing AWM has been EAB-driven mortality of *Fraxinus spp.*, but *Fraxinus* AWM has started declining as most individuals ≥10 cm have already died (Fig. `r mortality_by_genera`).
It is possible that increasing AWM will be maintained by trends of increasing oak mortality or by other non-indigenous insects or pathogen, which have arrived in the region in multiple waves over the past century [@anderson-teixeira_longterm_2021]. 
The ongoing emergence of new threats endangers other important canopy species: beech leaf disease, a non-indigenous pathogen lethal to *Fagus grandifolia* and first detected in 2012, is rapidly spreading throughout eastern US deciduous forests [@reed_distribution_2022].
Another non-indigenous pathogen, Sudden Oak Death, is currently confined to the Western US but is causing widespread mortality in multiple Oak species and poses huge risks if it were to successfully spread to this region [@conrad_threat_2020]. 
Thus, we consider it likely that oak decline, additional non-indigenous insects or pathogen species, and climate stressors will continue to elevate tree mortality at this site in the coming decades.
Similarly, the observed long-term declines at this site in AWP (Fig. `r loss_gain_net`b) and individual tree growth [@anderson-teixeira_joint_2022] remain poorly understood, hindering prediction of their future trends.
Overall, there remains large uncertainty as to the future trend in AGB at this site, but our analysis makes it clear that these trends have been and will continue to be substantially impacted by nuisance species.

Despite inherent limitations in extrapolating findings into the future and from a single site to a broader region, we find robust evidence suggesting a broader applicability of our results. 
Mixed-hardwood forests, akin to our study site, are typical of the mid-Atlantic region.
Furthermore, the widespread occurrence of vulnerable species (*e.g.* *Fraxinus* spp.) and continued spread of non-indigenous nuisance species (*e.g.* emerald ash borer, EAB) reinforces the relevance of our findings to the broader regional context.
EAB consistently induces near-total mortality in *Fraxinus* spp. once infected, to such an extent that five species in North America (including *Fraxinus americana*) are now critically endangered (Knight et al, 2013, IUCN Red list).
Similarly, temperate forests of eastern North America continue to be affected by a variety of non-indigenous pests and pathogens [@anderson-teixeira_longterm_2021; @aukema_historical_2010; @fei_biomass_2019], and emerging threats (e.g., beach leaf disease, sudden oak death) could also have profound impacts [@reed_distribution_2022;@conrad_threat_2020] on forests across the region.
<!--MOVED UP:
The ongoing emergence of new threats endangers other important canopy species: beech leaf disease, a non-indigenous pathogen lethal to *Fagus grandifolia* and first detected in 2012, is rapidly spreading throughout eastern US deciduous forests [@reed_distribution_2022].
Another non-indigenous pathogen, Sudden Oak Death, is currently confined to the Western US but is causing widespread mortality in multiple Oak species and poses huge risks if it were to successfully spread to this region [@conrad_threat_2020]. -->
Also paralleling our site, persistently elevated deer populations are reported throughout eastern US forests, and pressures from deer brousing have contributed to regneration debts of canopy species throughout the region [@miller_compounding_2019; @miller_overabundant_2023].
